---
title: 'STAT 4224 HW #1'
author: "Carlyle Morgan"
date: "9/22/2021"
output: pdf_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1.

## a.

\begin{table}[h!]
\begin{tabular}{|l|l|l|l|l|}
\hline
 & X=0 & X=1 & X=2 & X=3 \\ \hline
Y=0 & 1/12 & 1/4 & 3/20 & 1/60 \\ \hline
Y=1 & 1/60 & 3/20 & 1/4 & 1/12 \\ \hline
\end{tabular}
\end{table}


## b.

$E(Y|X=0) = \frac{1}{2}(1) + \frac{3}{10}(2) + \frac{1}{30}(3) = \frac{6}{5}$
$E(Y|X=1) = \frac{3}{10}(1) + \frac{1}{2}(2) + \frac{1}{6}(3) = \frac{9}{5}$
$E(Y) = (\frac{1}{4} + \frac{3}{20})(1) +(\frac{1}{4} + \frac{3}{20})(2) + (\frac{1}{60}+\frac{1}{12})(3) = \frac{3}{2}$

$E(E(Y|X)) = (\frac{6}{5} + \frac{9}{5})\frac{1}{2} = \frac{3}{2} =E(Y)$

## c.
$var(Y|X=0) = \frac{1}{2}(1) + \frac{3}{10}(4) + \frac{1}{30}(9) - (\frac{6}{5})^2 = \frac{14}{25}$

$var(Y|X=1) = \frac{3}{10}(1) + \frac{1}{2}(4) + \frac{1}{6}(9) - (\frac{9}{5})^2 = \frac{14}{25}$

$var(Y) = (\frac{1}{4} + \frac{3}{20})(1) +(\frac{1}{4} + \frac{3}{20})(4) + (\frac{1}{60}+\frac{1}{12})(9) - (\frac{3}{2})^2 = \frac{13}{20}$

$E(var(Y|X)) = (\frac{14}{25})(\frac{1}{2}) +  (\frac{14}{25})(\frac{1}{2}) = \frac{14}{25}$

$var(E(Y|X)) = (\frac{36}{25})(\frac{1}{2}) +  (\frac{81}{25})(\frac{1}{2}) - (\frac{3}{2})^2 = \frac{9}{100}$

Thus, $E(var(Y|X))+ var(E(Y|X)) = \frac{14}{25} + \frac{9}{100} = \frac{13}{20} = var(Y)$

## d. 
$p(x=1|y=2) = \frac{\frac{1}{4}}{\frac{3}{20}+\frac{1}{4}} = \frac{5}{8}$

# 2.

## a.
$p(x|y,z) = \frac{p(x,y,z)}{p(y,z)} = \frac{p(x,y,z)}{\int p(x',y,z)dx'}$

$= \frac{f(x,z)g(y,z)}{\int f(x',z) g(y,z) dx'}$

$= \frac{f(x,z)}{\int f(x',z)dx'}$

$=\frac{f(x,z)}{h^*(z)} = f^*(x,z)$

## b.

$p(y|x,z) = \frac{p(x,y,z)}{p(x,z)} = \frac{p(x,y,z)}{\int p(x,y',z)dy'}$

$= \frac{f(x,z)g(y,z)}{\int f(x,z) g(y',z) dy'}$

$= \frac{g(y,z)}{\int g(y',z)dy'}$

$=\frac{g(y,z)}{h^*(z)} = g^*(y,z)$

## c.

$p(x|z) = \int p(x|y,z)p(y|z)dy = f^*(x,z) \int p(y|z)dy =f^*(x,z)$

$p(y|z) = \int p(y|x,z)p(x|z)dy = g^*(y,z) \int p(x|z)dy =g^*(y,z)$


$\Rightarrow p(x,y|z) = p(x|z)p(y|x,z)$

$=f^*(x,z)g^*(y,z) = p(x|z)p(y|z)$

# 3.

## a.
$p(y_1,...y_n|\theta) = \prod_{i=1}^np(y_i|\theta) = \prod_{i=1}^n \theta^{y_i}(1-\theta)^{1-y_i} = \theta^{\sum y_i}(1-\theta)^{n-\sum y_i}$

$p(\sum_{i=1}^n y_i = y|\theta) = {n \choose y}\theta^{y}(1-\theta)^{n-y}$

## b.
```{r}
theta<-seq(0,1,.1)
likelihood<-dbinom(58,100,theta)
likelihood
plot(theta,likelihood, type = "l")
```

## c.
```{r}
likelihood<-dbinom(58,100,theta)
post<-likelihood/sum(likelihood)
plot(theta,post, type = "l")
```

## d.
```{r}
theta<-seq(0,1,.001)
likelihood<-dbinom(58,100,theta)
plot(theta,likelihood, type = "l")
```

## e.
```{r}
post<-dbeta(theta,58+1,100+1-58)
plot(theta,post,type="l")
```

All the graphs have the same pattern across theta $\in [0,1]$.

# 4.

## a.

$p(Child has Hh|Child has brown eyes, parents both have brown eyes)$

$=\frac{p(Child has Hh,Child has brown eyes, parents both have brown eyes)}{p(Child has brown eyes, parents both have brown eyes)}$

$=\frac{p(childhasHh,parentsbothHh)+p(childhasHh, one parent has Hh)+p(child has Hh, parents both HH)}{p(childhasbrowneyes,parentsbothHh)+p(childhasbrowneyes, one parent has Hh)+p(child has browneyes, parents both HH)}$

$=\frac{\frac{1}{2}4p^2(1-p)^2+\frac{1}{2}4p(1-p)^3+0(1-p)^4}{\frac{3}{4}4p^2(1-p)^2+4p(1-p)^3+(1-p)^4}$

$=\frac{2p}{1+2p}$

## b.

### i.

$p(Judi is Hh|Judy has brown eyes, Judy's parents have brown eyes, Judy's partner is Hh,p)$

$=\frac{\frac{2p}{1+2p}\frac{3}{4}}{\frac{2p}{1+2p}\frac{3}{4}+\frac{1}{1+2p}} = \frac{3p}{3p+2}$

$\hat p^2 = .16 \Rightarrow \hat p = .4$

$\Rightarrow$ The posterior probability is $\frac{3}{8}$

### ii.

The child must be of Hh for the grandchild to have blue eyes. Given $\hat p = .4$: 

$p(child is Hh|...) = \frac{3}{8}\frac{2}{3} + \frac{5}{8}\frac{1}{2} = \frac{9}{16}$

$p(grandchild is Hh|...) = \frac{9}{16}(\frac{1}{4}2p(1-p)+\frac{1}{2}p^2) = \frac{9}{80}$

